---
title: "DataExploration"
author: "Will"
date: "`r Sys.Date()`"
output: html_document
---

Notes on Variables: ALEGATN1 - Specific Malpractice Act or Omission Code
ALEGATN2 - First used 1/31/2004. Second A&O Code OUTCOME - Severity of
Malpractice Injury - is available for REPTYPE P (payment reports
beginning 1/31/2004) MALYEAR1 - Contains some errors (e.g., 3999).
Beginning year of A&O MALYEAR2 - Contains some errors. End year of A&O.
If same as MALYEAR1, may be blank PAYMENT - recorded as midpoint of an
interval. Amt of payment that led to filing of medical malpractice
report TOTALPMT - Generally the same as Payment. May be \> PAYMENT if
other payments are made by the reporting entity for this practitioner in
this case. NUMBPRSN - Number of practitioner involved in case. Note that
payment refers to the specific practitioner. PAYTYPE - Judgement or
settlement. Anything other than J is recommended to be considered a
Settlement PYRRLTNS - How is paying entity related to the practitioner.
Insurance Co, etc. PTTYPE - Inpatient, outpatient, both, unknown
AACLASS1 - not relevant to payments BASISCD# - basis code for Adverse
action AALENTYP - Indefinite, Permanent, or Specified penalty length
AALENGTH - Length of Specified penalty AAEFYEAR - Effective Year of
adverse action AASIGYR - Signature year of adverse action TYPE - Type of
reporting entity PRACTNUM - For linking different records for the same
practitioner ACCRRPTS - Total number of accredidation reports NPMALRPT -
Total Malpractice payment reports for a practitioner NPLICRPT -
Subject's Number of Clinical Privileges or Panel Member Reports
submitted under Title IV NPPSMRPT - Subject's Number of Professional
Society Membership Reports submitted under Title IV NPDEARPT - Subject's
Number of DEA Reports submitted under Title IV NPEXCRPT - Subject's
Number of Exclusion Reports submitted under Title IV and/or Section 1921
NPGARPT - Subject's Number of Government Administrative Reports
submitted under Section 1921. NPCTMRPT - Subject's Number of Contract
Termination Reports submitted under Section 1921 FUNDPYMT - 0
Malpractice Payment Made by an Insurance Company or Self-Insured Entity
1 Malpractice Payment Made by a State Fund




# change from haven type to factor

# <https://stackoverflow.com/questions/67333840/error-in-usemethodrescale-no-applicable-method-for-rescale-applied-to-an>



Insurance companies are the main insurer.

```{r RECTYPE}
data %>% count(RECTYPE) %>% dplyr::arrange(desc(n))
data %>% dplyr::count(RECTYPE)
data %>% group_by(RECTYPE) %>% summarise_at(vars(PAYMENT), list(mean_pmt = mean))
summary(data[data$RECTYPE == "M","PAYMENT"])
data[data$RECTYPE == "M" & is.na(data$PAYMENT), ]
data[data$REPTYPE == 101, "RECTYPE"] %>% group_by(RECTYPE) %>% count()
```

The NA values are all for REPTYPE 101, and mostly in NY

There are some NA values for PAYMENT in records of type "M". Since M is
the old payment type, this should be an error.

There are around 500k payments, with the rest being adverse actions

```{r REPTYPE}
data %>% count(REPTYPE) %>% dplyr::arrange(desc(n))
```

302 - State Licensure or Certification Action (Indiv, CAAR, 11/22/99 and
later) is the most common report type. Payments under type 101 and 102
are still substantial and there is considerable opportunity to model
payments. Grouping by broader categories of report type (licensure,
professional society, etc.) may yield useful insights, though the
breakdown by RECTYPE with new and old combined may be the best route.

```{r ORIGYEAR}
orig_yr_cts = data %>% count(ORIGYEAR) %>% dplyr::arrange(desc(n))
plot(x=orig_yr_cts$ORIGYEAR, y=orig_yr_cts$n)
```

There is an upwards trend in the mean and variance for the number of
filed reports.

```{r WORKSTAT}
data %>% count(WORKSTAT) %>% arrange(desc(n))
```

```{r PAYMENT_REPORT_TYPES}
data %>% group_by(PAYMENT_DUMMY, REPTYPE) %>% count() %>% filter(PAYMENT_DUMMY == 1)
```

REPTYPEs for Payments are all 101 and 102

```{r PAYMENT_TREND}
payment_trend = data %>% group_by(ORIGYEAR) %>% count(PAYMENT_DUMMY)
head(payment_trend)
ggplot(payment_trend, aes(ORIGYEAR, n, colour = as.factor(PAYMENT_DUMMY))) + geom_point()
```

Above, we see that payments are decreasing while non-payment reports are dramatically increasing. Narrowing in on payments, lets look for trends in the total payments.

