---
title: "LinReg_Project"
author: "Will Mohr"
date: "2024-11-12"
output: pdf_document
---

```{r import_data, echo=FALSE}
library(haven)
library(dplyr)

data = read_por("./NPDB2404.POR")
summary(data)
print(head(data[complete.cases(data$PAYMENT),],3))
```

```{r}
data %>% count(RECTYPE) %>% dplyr::arrange(desc(n))

data$REC_PMT = ifelse()
help(ifelse)
```
There are around 500k payments, with the rest being adverse actions
```{r}
data %>% count(REPTYPE) %>% dplyr::arrange(desc(n))
```
302 - State Licensure or Certification Action (Indiv, CAAR, 11/22/99 and later) is the most common report type.

Payments under type 101 and 102 are still substantial and there is considerable opportunity to model payments.

Grouping by broader categories of report type (licensure, professional society, etc.)
may yield useful insights, though the breakdown by RECTYPE with new and old combined may be the best route.

```{r}
orig_yr_cts = data %>% count(ORIGYEAR) %>% dplyr::arrange(desc(n))

plot(x=orig_yr_cts$ORIGYEAR, y=orig_yr_cts$n)
```

```{r}
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
